---
title: "`aswp`: An opinionated LaTeX template for R Markdown working papers"
subtitle: A Short Introduction
author:
  - name: András Svraka
    affiliation: |
      This is from the optional `affiliation` field. It's just an
      unnumbered footnote for each author. Besides affiliation, you
      can put e.g. contact information here.
date: \today
thanks: |
  This is from the optional `thanks` field. Use this to add further notes that do
  not not belong to any specific author. I'm standing on the shoulders
  of giants.
abstract: |
  This short vignette gives an introduction to the `aswp` R Markdown
  template. This template provides nice typography for a research
  paper and offers a simple way to add paragraph-long notes to
  figures.
bibliography: aswp.bib
keywords: [R Markdown, Reproducible research, Typography]
jel: [C88, Y20]
output: aswp::aswp

# The following section is required for vignettes. Don't use it for
# other documents.
vignette: >
  %\VignetteIndexEntry{aswp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(aswp)
```

# First steps

This package provides an `aswp()` function which is an R Markdown document format creator. To use it, you need to add `aswp::aswp` to the `output` field in the YAML metadata block of your R Markdown file. The recommended way to use this template is however by

```{r, eval = FALSE}
rmarkdown::draft("paper.Rmd", template = "aswp", package = "aswp")
```

Or use your editor's interface to this function:

  * In RStudio, open a new R Markdown document (**File** → **New File** and choose **R Markdown**) and select this template ("Working Paper") under **From Template**.
  * In Emacs, use `M-x poly-r-rmarkdown-create-from-template` with [Polymode](http://polymode.github.io/), which is also available from the iESS menu.

This provides a basic skeleton for the most important, and non-obvious features and helps setting up a more complex document with references and appendices which are often used in research papers.

Then knit and compile like any other R Markdown document and you'll get a nicely formatted PDF. [TinyTeX](https://yihui.org/tinytex/) should automatically take care of the required \LaTeX{} packages.

# Metadata fields

The basic YAML metadata fields for `title`, `subtitle`, `abstract`, `data` and `thanks` are available and behave similarly to the standard R Markdown template used in `rmarkdown::pdf_document()`. The `keywords` fields is also available and keywords are displayed after the abstract, not just added to the PDF's metadata fields. Additionally, a list of [JEL Codes](https://www.aeaweb.org/econlit/jelCodes.php?view=jel) can be added in the same way using the `jel` field. These are also displayed after the abstract.

The `author` field works differently than in other templates, you need to add a `name` label, and an optional `affiliation` can be used to add affiliations and contact information.

``` yaml
author:
  - name: First Author
    affiliation: Institute of Whatever, <name@whatever.gov>
  - name: Second Author
    affiliation: Company, Inc., <name@company.com>
```

# Citations

Add a `bibliography` field to the YAML frontmatter with the name of your bibliography file and cite works in the text using Pandoc's standard [citation forms](https://www.pandoc.org/MANUAL.html#citations), e.g. `@Xie2019` for @Xie2019. Both Bib\LaTeX{} and Pandoc's default `pandoc-citeproc` can be used for citation rendering but `pandoc-citeproc` is recommended. It is the only way currently to put a "References" section before the appendices. Unfortunately Pandoc is not aware of \LaTeX{}'s concept of appendices so you need to do the following incantations to add all sections to the paper in the right order, and with the right numbering.[^1]

[^1]: I'm not aware of similar hacks for Bib\LaTeX{}.

``` markdown
This is the last sentence of the paper.

# References {-}

<div id="refs"></div>

\appendix
\clearpage

# Appendix

Start appendix here.
```

If you start your document with a template from this package (`rmarkdown::draft()`), references and an appendix section are already set up.

# R Markdown {#sec:rmarkdown}

All the usual R Markdown features work but there are some extra functionalities.

This package builds on [**bookdown**](https://github.com/rstudio/bookdown) to produce a single long-form document. This means chapters are not available but you can use cross-references with the `\@ref(...)` syntax for [equations](https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html#equations) and [text references](https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html#text-references): See Section \@ref(sec:rmarkdown).

## Figure notes

It is often useful to add a paragraph below a figure with notes, or references. This allows self-contained figures with short, concise captions yet you can add more context. Standard R Markdown and **knitr** doesn't support this for floating environments, like \LaTeX{} figures and tables.

For this feature you need to load this package at the top your R Markdown document: `library(aswp)`. This will set up **knitr** hooks to insert `floatfoot` environments from the `floatrow` \LaTeX{} package by using `floatfoot` chunk options. See Figure \@ref(fig:plot-note) for a simple example.

````markdown
`r ''````{r chunk-label, fig.cap = "A figure caption.",
floatfoot = "A figure note."}
plot(cars)
```
````

```{r plot-note, fig.cap = "A figure caption.", floatfoot = "A figure note.", fig.height = 3.5, echo = FALSE}
par(mar = c(4, 4, .1, .1))
plot(cars)
```

It is however quite inconvenient to write longer sentences, or even full paragraphs as **knitr** chunk options. And you'd have to write raw \LaTeX{} code and escape all the backslashes. Fortunately **bookdown** has a nice featre where you can use text references anywhere. Define them:

```markdown
(ref:foo) Define a text reference **here**.
```

Then you can use `(ref:foo)` as caption, or note. Both the caption and the note of Figure \@ref(fig:text-ref) were written in Markdown.

```{r text-ref, fig.cap='(ref:text-ref-cap)', floatfoot = '(ref:text-ref-note)', fig.height = 3.5, echo = FALSE}
par(mar = c(4, 4, .1, .1))
plot(cars)

n <- nrow(cars)
```

(ref:text-ref-cap) A scatterplot of the data `cars` using **base** R graphics.
(ref:text-ref-note) **Note:** This my note from a *text reference*. It is a \emph{very} long note that should overflow into multiple rows. This is all possible thanks to **bookdown** [@Xie2019]. Figure \@ref(fig:plot-note) is much simpler. Let's mix some maths and R code, this dataset has $N = `r n`$ observations. Unfortunately this note [cannot be split into separate paragraphs](https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html#text-references) but that's probably OK.

# References {-}

<div id="refs"></div>

\appendix
\clearpage

# Appendix

This is the appendix, which is numbered differently. You can even make citations [@Xie2019]


## Subsection
